Algorithmic detection of conserved quantities of finite-difference schemes for partial differential equations

### 2023-05-07 Papers/Paper ACM/paper.tex
paper changes

\usepackage{amsmath}
%\usepackage{amssymb}
\usepackage{amsthm}
\usepackage[latin1]{inputenc}
\usepackage{hyperref}
\usepackage{dsfont}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage{ifthen}
\usepackage{mmacells}
\usepackage[invisible]{MLaTex}

@@

\section{Introduction}
%\label{intro}
Many partial differential equations (PDEs) admit integral quantities, that are conserved in time. 
For example, the advection equation preserves energy and the heat equation conserves mass. 
When approximating PDEs by a finite-difference scheme, the question arises whether such quantities are conserved by the scheme. 
Such information can be crucial to estimate whether a scheme approximates a PDE accurately and to determine its stability.\\
Computations for determining conservation can get rather tedious.
Hence, automating them in computer algebra systems is desirable.\\
Conserved quantities correspond to conservation laws, admitted by the PDE. A conservation law \cite{cheviakov20} is an equation of the form

@@

	\caption{{\sc time-explicit DiscreteConservedQ}}
	\label{alg: discreteconservedqexplicit}
\end{algorithm}
Note, that the result of Step 2 exists only at time $t$ because all instances of $t+1$ have been replaced. Thus, we do not need any computations of difference ideals or Gr{\"o}bner basis.
\begin{example}
	\label{timeexplicitheateqalgo}
	We check for conservation of $\sum_nu(n,t)$ under the scheme
	\begin{equation*}
		u(n,t+1)-u(n,t)-(u(n+1,t)-2u(n,t)+u(n-1,t)).
	\end{equation*}
	The discrete time derivative (Step 1) is
	\begin{gather*}
		\sum_{n\in\Ii}u(n,t+1)-u(n,t).
	\end{gather*}
	Replace $u(n,t+1)$ by the right-hand side (Step 2) to get
	\begin{gather*}
		\sum_{n\in\Ii}u(n,t)+(u(n+1,t)-2u(n,t)+u(n-1,t))-u(n,t).
	\end{gather*}
	Then, we compute the discrete partial variational derivative (Step 3), which equals zero. Hence (Step 4), we have conservation.
\end{example}
\begin{example}
	\label{heatex1}
	We verify the result from Example \ref{timeexplicitheateqalgo}, using our implementation in {\sc Mathematica} of Algorithm \ref{alg: discreteconservedqexplicit}.
	\begin{EXE}
		LABEL="rhs1"
		(*CODE*)
		variables = Association["indVars"->{n},"depVars"->{u},"rhs"->{u[n]+(u[n+1]-2*u[n]+u[n-1])}]
		expression = u[n]
		DiscreteConservedQOperator[variables][expression]
		(*CODE*)
		PUNCT={,.}
	\end{EXE}
	\begin{small}
		\input{executables/rhs1ctt.ed.tex}
		\input{executables/rhs1ctt.result.ed.tex}
	\end{small}
\end{example}
\section{The general case}
In this section, we present Algorithm \ref{alg: discreteconservedqimplicit} that deals with general, not necessarily time-explicit, schemes. We explain its steps in detail below and demonstrate the algorithm in Example \ref{heateqalgoimplicit}.\\
%In this section we explain our algorithm, that checks if a given quantity is preserved in time under a given scheme in detail.
\begin{algorithm}
	\DontPrintSemicolon
	\KwIn{$\mathcal{J}\in\mathcal{F}(\mathcal{P}(\Ii\times\Nn,\Rr^m),\mathcal{D}(\Nn))$ and $\{H^1,\ldots,H^i\}\subset\mathcal{S}(\mathcal{P}(\Ii\times\Nn,\Rr^m),\mathcal{P}(\Ii\times\Nn,\Rr))$ }
	\KwOut{ {\sc True}, if $\mathcal{J}$ is conserved in time, {\sc False} otherwise }

>>>>

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage[latin1]{inputenc}
\usepackage{hyperref}
\usepackage{dsfont}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage{ifthen}
\usepackage{mmacells}
\usepackage[invisible]{MLaTex}

@@

\section{Introduction}
%\label{intro}
Many partial differential equations (PDEs) admit integral quantities, that are conserved in time. 
For example, the advection equation preserves energy and the heat equation conserves mass. 
When approximating PDEs by a finite-difference scheme, the question arises whether such quantities are conserved by the scheme. 
Such information can be crucial to estimate whether a scheme approximates a PDE accurately and to determine its stability.\\
Computations for determining conservation can get rather tedious.
Hence, automating them in computer algebra systems is desirable.\\
Conserved quantities correspond to conservation laws, admitted by the PDE. A conservation law \cite{cheviakov20} is an equation of the form

@@

	\caption{{\sc time-explicit DiscreteConservedQ}}
	\label{alg: discreteconservedqexplicit}
\end{algorithm}
Note, that the result of Step 2 exists only at time $t$ because all instances of $t+1$ have been replaced. Thus, we do not need any computations of difference ideals or Gr{\"o}bner basis.
\begin{example}
	\label{timeexplicitheateqalgo}
	We check for conservation of $\sum_nu(n,t)$ under the scheme
	\begin{equation*}
		u(n,t+1)-u(n,t)-(u(n+1,t)-2u(n,t)+u(n-1,t)).
	\end{equation*}
	The discrete time derivative (Step 1) is
	\begin{gather*}
		\sum_{n\in\Ii}u(n,t+1)-u(n,t).
	\end{gather*}
	Replace $u(n,t+1)$ by the right-hand side (Step 2) to get
	\begin{gather*}
		\sum_{n\in\Ii}u(n,t)+(u(n+1,t)-2u(n,t)+u(n-1,t))-u(n,t).
	\end{gather*}
	Then, we compute the discrete partial variational derivative (Step 3), which equals zero. Hence (Step 4), we have conservation.
\end{example}
\begin{example}
	\label{heatex1}
	We verify the result from Example \ref{timeexplicitheateqalgo}, using our implementation in {\sc Mathematica} of Algorithm \ref{alg: discreteconservedqexplicit}.
	\begin{EXE}
		LABEL="rhs1"
		(*CODE*)
		variables = Association["indVars"->{n},"depVars"->{u},"rhs"->{u[n]+(u[n+1]-2*u[n]+u[n-1])}]
		expression = u[n]
		DiscreteConservedQOperator[variables][expression]
		(*CODE*)
		PUNCT={,.}
	\end{EXE}
	\begin{small}
		\input{executables/rhs1ctt.ed.tex}
		\input{executables/rhs1ctt.result.ed.tex}
	\end{small}
\end{example}
\section{The general case}
\label{secgen}
In this section, we present Algorithm \ref{alg: discreteconservedqimplicit} that deals with general, not necessarily time-explicit, schemes. We explain its steps in detail below and demonstrate the algorithm in Example \ref{heateqalgoimplicit}.\\
%In this section we explain our algorithm, that checks if a given quantity is preserved in time under a given scheme in detail.
\begin{algorithm}
	\DontPrintSemicolon
	\KwIn{$\mathcal{J}\in\mathcal{F}(\mathcal{P}(\Ii\times\Nn,\Rr^m),\mathcal{D}(\Nn))$ and $\{H^1,\ldots,H^i\}\subset\mathcal{S}(\mathcal{P}(\Ii\times\Nn,\Rr^m),\mathcal{P}(\Ii\times\Nn,\Rr))$ }
	\KwOut{ {\sc True}, if $\mathcal{J}$ is conserved in time, {\sc False} otherwise }
